---
title: "OECD regional inequality 2018"
output: github_document
---

A primer on regional inequality and data

Set up, load in libraries

```{r}
#path <- your.path.to.our.repo
#setwd(path)

#libs
libs <- c("tidyverse", "dtplyr", "data.table", "OECD", "knitr", "countrycode")
lapply(libs, require, character.only=T)


```


The OECD data we're going to work with is available here: https://stats.oecd.org/Index.aspx?DataSetCode=REGION_DEMOGR#

These files are massive and complex. We've downloaded the data in advance and made it available in the repo's "inputs" folder.

Let's read in the file and inspect the vars

```{r}
reg.dat <- fread("inputs/oecd_raw_dat.csv")

#Inspect the data, by code and descriptor var pairs
reg.dat[, .N, .(TL, Territory.Level.and.Typology)] #TL level
reg.dat[, .N, .(REG_ID, Region)] #Region
reg.dat[, .N, .(SERIES, SNA.Classification)] #Series
reg.dat[, .N, .(VAR, Indicator)] #Indicator
reg.dat[, .N, .(MEAS, Measure)] #Measure
reg.dat[, .N, .(POS, Position)] #Position
reg.dat[, .N, .(TIME, Year)] %>% head() #time
reg.dat[, .N, .(Unit.Code, Unit)] #units
reg.dat[, .N, .(PowerCode.Code, PowerCode)] #powercode
reg.dat[, .N, .(Reference.Period.Code, Reference.Period)] #ref period
reg.dat[, .N, .(Flag.Codes, Flags)] #flags

#We also need a names lookup file that we created
oecd.names <- fread("inputs/oecd_names_lookup.csv")

```


First let's explore all the measures in our dataset, for UK TL3 (small) areas alone

```{r}
#---------- Which measure is best?  ----------------

#Check out the GDP measures
reg.dat[grepl("3", TL),] %>%
  .[, .(REG_ID, Region, VAR, Indicator, MEAS, Measure, Year, Value)] %>% .[, .N, .(VAR, MEAS, Measure)]

#Total GDP at TL3 (d1)
reg.dat[VAR == "GDP" & MEAS == "REAL_PPP" & grepl("3", TL),] %>%
  .[grepl("UK", REG_ID) & Year == 2016,] %>%
  .[, .(REG_ID, Region, Indicator, Measure, Year, Value)] %>%
  arrange(-Value) -> d1 #%>% kable(.)
head(d1) %>% kable

#GDP per person at TL3
reg.dat[VAR == "GDP" & MEAS == "PC_REAL_PPP" & grepl("3", TL),] %>%
  .[grepl("UK", REG_ID) & Year == 2016,] %>%
  .[, .(REG_ID, Region, Indicator, Measure, Year, Value)] %>%
  arrange(-Value) -> d2 #%>% kable(.)
head(d2) %>% kable

#Population by region (implicit)
full_join(d1, d2, by = c("REG_ID", "Region")) %>%
  .[, `:=`(Measure = "Population", Indicator = "Pop", Year = 2016, Value = Value.x * 10^6 / Value.y)] %>%
  .[, .(REG_ID, Region, Indicator, Measure, Year, Value)] %>%
  arrange(Value) -> d3
head(d3) %>% kable

#Work-based employment
reg.dat[VAR == "EMP_IND_TOTAL" & grepl("3", TL),] %>%
  .[grepl("UK", REG_ID) & Year == 2016,] %>%
  .[, .(REG_ID, Region, Indicator, Measure, Year, Value)] %>%
  arrange(-Value) -> d4
head(d4) %>% kable

#GDP per person employed
full_join(d1, d4, by = c("REG_ID", "Region")) %>%
  .[, `:=`(Measure = "GDP per person employed", Indicator = "GDP per person", Year = 2016, Value = Value.x * 10^6 / Value.y)] %>%
  .[, .(REG_ID, Region, Indicator, Measure, Year, Value)] %>%
  arrange(-Value) -> d5
head(d5) %>% kable

#Regional Gross Value Added (GDP ex. subsidies)
reg.dat[VAR == "GVA_IND_TOTAL" & MEAS == "PW_REAL_PPP" & grepl("3", TL),] %>%
  .[grepl("UK", REG_ID) & Year == 2016,] %>%
  .[, .(REG_ID, Region, Indicator, Measure, Year, Value)] %>%
  arrange(-Value) -> d6 #%>% kable(.)
head(d6) %>% kable
#n.b not available for all TL3 regions

rbindlist(l = list(d1, d2, d3, d4, d5, d6), fill=T) -> dat.uk
head(dat.uk) %>% kable

```

With this data gathered, let's see how things look in Camden & City of London versus the Isle of Angelsey: 

```{r}
#Grab Camden & Anglesey
dat.uk[grepl("Camden", Region),] %>% kable
dat.uk[grepl("Torbay", Region),] %>% kable

#Comparing measures
dat.uk[REG_ID %in% c("UKL11", "UKI31"),] %>%
  dcast(Measure ~ Region, value.var = "Value") %>%
  rename(Camden.City=2, Anglesey=3) %>%
  mutate(mutiple = round(Camden.City / Anglesey, 3)) %>% kable(.)

```

The table above shows descriptive the values for each var, and the multiples between them. 

Now, let's produce the same variables for every OECD country. 

First we'll being with GDP per person: 


```{r}

#--------- Get GDP per person for different geographic areas ---------

#National 
reg.dat[VAR == "GDP" & MEAS == "PC_REAL_PPP" & TL == 1,] %>%
  .[!REG_ID %in% c("OECD"), .(REG_ID, Region, Measure, Year, Value)] %>%
  left_join(., rename(oecd.names, REG_ID = oecd.nat.code), by = "REG_ID") %>%
  .[, .(iso3c, REG_ID, Region, Measure, Year, Value)] -> dat2.nat

#GDP per person at TL2 
reg.dat[VAR == "GDP" & MEAS == "PC_REAL_PPP" & grepl("2", TL),] %>%
  .[, .(REG_ID, Region, Measure, Year, Value)] %>%
  .[, oecd.imp.code := substr(REG_ID, 1, 2)] %>%
  left_join(., oecd.names, by = "oecd.imp.code") -> dat2.tl2

#GDP per person at TL3 
reg.dat[VAR == "GDP" & MEAS == "PC_REAL_PPP" & grepl("3", TL),] %>%
  .[, .(REG_ID, Region, Measure, Year, Value)] %>%
  .[, oecd.imp.code := substr(REG_ID, 1, 2)] %>%
  left_join(., oecd.names, by = "oecd.imp.code") -> dat2.tl3

```

With the main data isolated, dat2, for national, tl2 and tl3, let's contstruct the plot data

First, the richest areas in each TL3 (TL2s for USA)

```{r}

#--------- Richest regions ---------

print("Richest regions, latest data")
dat2.tl3[, .SD[which.max(Value)], by=.(iso3c, Year)] %>%
  .[, .SD[which.max(Year)], by=.(iso3c)] %>% 
  .[, geo := "richest.region"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d2.richest.latest

#Adding richest region in USA (TL2) to TL3 regions
dat2.tl2[, .SD[which.max(Value)], by=.(iso3c, Year)] %>%
  .[, .SD[which.max(Year)], by=.(iso3c)] %>% 
  .[, geo := "richest.region"] %>%
  .[iso3c == "USA", .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  bind_rows(., d2.richest.latest) %>% 
  arrange(iso3c) -> d2.richest.latest
d2.richest.latest %>% kable

print("Richest regions in 2000, or earliest avail.")
dat2.tl3[REG_ID %in% d2.richest.latest$REG_ID,] %>%
  .[Year >= 2000, .SD[which.min(Year)], by=.(iso3c)] %>%
  .[, geo := "richest.region"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d2.richest.earliest

#Adding in richest region in USA (TL2) in 2000 to TL3 regions in 2000
dat2.tl2[REG_ID %in% d2.richest.latest$REG_ID,] %>%
  .[Year >= 2000, .SD[which.min(Year)], by=.(iso3c)] %>%
  .[, geo := "richest.region"] %>%
  .[iso3c == "USA", .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  bind_rows(., d2.richest.earliest) %>% 
  arrange(iso3c) -> d2.richest.earliest
d2.richest.earliest %>% kable

```



And repeat the steps above for the poorest regions

```{r}

#--------- Poorest regions ------------

print("Poorest regions, latest data")
dat2.tl3[!REG_ID %in% c("FRY50", "FRY30"), ] %>% #removing France's overseas departments
  .[, .SD[which.min(Value)], by=.(iso3c, Year)] %>% 
  .[, .SD[which.max(Year)], by=.(iso3c)] %>% 
  .[, geo := "poorest.region"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d2.poorest.latest

#Adding richest region in USA (TL2) to TL3 regions
dat2.tl2[, .SD[which.min(Value)], by=.(iso3c, Year)] %>%
  .[, .SD[which.max(Year)], by=.(iso3c)] %>% 
  .[, geo := "poorest.region"] %>%
  .[iso3c == "USA", .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  bind_rows(., d2.poorest.latest) %>% 
  arrange(iso3c) -> d2.poorest.latest
d2.poorest.latest %>% kable

print("Poorest regions in 2000, or earliest avail")
dat2.tl3[REG_ID %in% d2.poorest.latest$REG_ID,] %>%
  .[Year >= 2000, .SD[which.min(Year)], by=.(iso3c)] %>%
  .[, geo := "poorest.region"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d2.poorest.earliest

#Adding in poorest region in USA (TL2) in 2000 to TL3 regions in 2000
dat2.tl2[REG_ID %in% d2.poorest.latest$REG_ID,] %>%
  .[Year >= 2000, .SD[which.min(Year)], by=.(iso3c)] %>%
  .[, geo := "poorest.region"] %>%
  .[iso3c == "USA", .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  bind_rows(., d2.poorest.earliest) %>% 
  arrange(iso3c) -> d2.poorest.earliest
d2.poorest.earliest %>% kable


```


And then the national averages, being careful to match the same years above: 

```{r}

#--------- National averages (matching years with richest and poorest above) ---------

print("Nat average, match year to latest year in richest/poorest")
dat2.nat[iso3c %in% d2.richest.latest$iso3c,]  %>%
  inner_join(., d2.richest.latest[,.(iso3c, Year)], by = c("iso3c", "Year")) %>%
  .[, geo := "national.av"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d2.nat.latest
d2.nat.latest %>% kable

print("Nat average, 2000 or earliest avail.")
dat2.nat[iso3c %in% d2.richest.earliest$iso3c,]  %>%
  inner_join(., d2.richest.earliest[,.(iso3c, Year)], by = c("iso3c", "Year")) %>%
  .[, geo := "national.av"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d2.nat.earliest
d2.nat.earliest %>% kable

```


Now gather the data together (bind) and plot it (roughly)

```{r}

#Bind data
rbindlist(l = list(d2.nat.latest, d2.nat.earliest, 
                   d2.richest.latest, d2.richest.earliest, 
                   d2.poorest.latest, d2.poorest.earliest)) %>%
  left_join(., oecd.names, by = "iso3c") %>%
  dcast(country + iso3c + Year ~ geo, value.var = "Value") %>%
  mutate(indx.poorest.region = poorest.region / national.av * 100, 
         indx.richest.region = richest.region / national.av * 100) -> rich.poor.1
rich.poor.1 %>% kable

#The countries in our original chart
plot.iso <- c("GBR", "DEU", "USA", "FRA", "KOR", "ITA", "JPA", "ESP", "SWE")

#Plot it
rich.poor.1 %>% 
  filter(iso3c %in% plot.iso) %>%
  select(country, Year, poorest=indx.poorest.region, richest=indx.richest.region) %>% 
  melt(id.vars = c("country", "Year"), variable.name = "geo") -> plot.dat.1
plot.dat.1 %>% kable
write_csv(plot.dat.1, "outputs/plot.dat.1.csv")
  
ggplot(plot.dat.1, aes(x=value, y=1, colour=Year)) + geom_point() + geom_text(aes(label=geo, vjust=-1)) +
  facet_wrap(~country, ncol=1) + xlim(0, 1200) + theme_minimal() +
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
  theme(legend.position="top") + ggtitle("Richest v poorest regions") -> plot1
print(plot1)
```


Now, as the Medium piece discusses while in context of the original Economist piece
discussing regional inequality (https://www.economist.com/briefing/2017/10/21/globalisation-has-marginalised-many-regions-in-the-rich-world) 
the measure is useful. It has since been taken out of context. 

So, let's introduce an alternative measure, keeping the richest and poorest regions (for ease of communication) but dividing a region's GDP not by 
its population but the number of employees working there. 

```{r}

#------- Total GDP ----------

#National 
reg.dat[VAR == "GDP" & MEAS == "REAL_PPP" & TL == 1,] %>%
  .[!REG_ID %in% c("OECD"), .(REG_ID, Region, Measure, Year, Value)] %>%
  left_join(., rename(oecd.names, REG_ID = oecd.nat.code), by = "REG_ID") %>%
  .[, .(iso3c, REG_ID, Region, Measure, Year, Value)] -> dat1.nat

#TL2 
reg.dat[VAR == "GDP" & MEAS == "REAL_PPP" & grepl("2", TL),] %>%
  .[, .(REG_ID, Region, Measure, Year, Value)] %>%
  .[, oecd.imp.code := substr(REG_ID, 1, 2)] %>%
  left_join(., oecd.names, by = "oecd.imp.code") -> dat1.tl2

#TL3 
reg.dat[VAR == "GDP" & MEAS == "REAL_PPP" & grepl("3", TL),] %>%
  .[, .(REG_ID, Region, Measure, Year, Value)] %>%
  .[, oecd.imp.code := substr(REG_ID, 1, 2)] %>%
  left_join(., oecd.names, by = "oecd.imp.code") -> dat1.tl3


#------- Work-based employment ---------

#National 
reg.dat[VAR == "EMP_IND_TOTAL" & TL == 1,] %>%
  .[!REG_ID %in% c("OECD"), .(REG_ID, Region, Measure, Year, Value)] %>%
  left_join(., rename(oecd.names, REG_ID = oecd.nat.code), by = "REG_ID") %>%
  .[, .(iso3c, REG_ID, Region, Measure, Year, Value)] -> dat4.nat

#TL2 
reg.dat[VAR == "EMP_IND_TOTAL" & grepl("2", TL),] %>%
  .[, .(REG_ID, Region, Measure, Year, Value)] %>%
  .[, oecd.imp.code := substr(REG_ID, 1, 2)] %>%
  left_join(., oecd.names, by = "oecd.imp.code") -> dat4.tl2

#TL3 
reg.dat[VAR == "EMP_IND_TOTAL" & grepl("3", TL),] %>%
  .[, .(REG_ID, Region, Measure, Year, Value)] %>%
  .[, oecd.imp.code := substr(REG_ID, 1, 2)] %>%
  left_join(., oecd.names, by = "oecd.imp.code") -> dat4.tl3


#------ GDP per person employed (implicit) --------

#National 
full_join(dat1.nat, dat4.nat, by = c("iso3c", "REG_ID", "Region", "Year")) %>%
  .[, `:=`(Measure = "GDP per person employed", Value = Value.x * 10^6 / Value.y)] %>%
  .[!is.na(Value), .(iso3c, REG_ID, Region, Measure, Year, Value)]  -> dat5.nat

#TL2
full_join(dat1.tl2, dat4.tl2, by = c("iso3c", "REG_ID", "Region", "Year")) %>%
  .[, `:=`(Measure = "GDP per person employed", Value = Value.x * 10^6 / Value.y)] %>%
  .[!is.na(Value), .(iso3c, REG_ID, Region, Measure, Year, Value)] -> dat5.tl2

#TL3
full_join(dat1.tl3, dat4.tl3, by = c("iso3c", "REG_ID", "Region", "Year")) %>%
  .[, `:=`(Measure = "GDP per person employed", Value = Value.x * 10^6 / Value.y)] %>%
  .[!is.na(Value), .(iso3c, REG_ID, Region, Measure, Year, Value)] -> dat5.tl3


```

Now we have the data we need, we'll breeze through the next steps as they repeat those for the GDP per person measure above. 

```{r}

#------- Richest regions, GDP per person --------- 

#Richest regions, latest data
dat5.tl3[, .SD[which.max(Value)], by=.(iso3c, Year)] %>%
  .[, .SD[which.max(Year)], by=.(iso3c)] %>% 
  .[, geo := "richest.region"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d5.richest.latest

#Adding richest region in USA (TL2) to TL3 regions
dat5.tl2[, .SD[which.max(Value)], by=.(iso3c, Year)] %>%
  .[, .SD[which.max(Year)], by=.(iso3c)] %>% 
  .[, geo := "richest.region"] %>%
  .[iso3c == "USA", .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  bind_rows(., d5.richest.latest) %>% 
  arrange(iso3c) -> d5.richest.latest
d5.richest.latest %>% kable

#Richest regions in 2000, or earliest avail. 
dat5.tl3[REG_ID %in% d5.richest.latest$REG_ID,] %>%
  .[Year >= 2000, .SD[which.min(Year)], by=.(iso3c)] %>%
  .[, geo := "richest.region"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d5.richest.earliest

#Adding in richest region in USA (TL2) in 2000 to TL3 regions in 2000
dat5.tl2[REG_ID %in% d5.richest.latest$REG_ID,] %>%
  .[Year >= 2000, .SD[which.min(Year)], by=.(iso3c)] %>%
  .[, geo := "richest.region"] %>%
  .[iso3c == "USA", .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  bind_rows(., d5.richest.earliest) %>% 
  arrange(iso3c) -> d5.richest.earliest
d5.richest.earliest %>% kable


#--------- Poorest regions, GDP per person employed ------------

#Poorest regions, latest data
dat5.tl3[!REG_ID %in% c("FRY50", "FRY30"), ] %>% #removing France's overseas departments
  .[, .SD[which.min(Value)], by=.(iso3c, Year)] %>% 
  .[, .SD[which.max(Year)], by=.(iso3c)] %>% 
  .[, geo := "poorest.region"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d5.poorest.latest

#Adding richest region in USA (TL2) to TL3 regions
dat5.tl2[, .SD[which.min(Value)], by=.(iso3c, Year)] %>%
  .[, .SD[which.max(Year)], by=.(iso3c)] %>% 
  .[, geo := "poorest.region"] %>%
  .[iso3c == "USA", .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  bind_rows(., d5.poorest.latest) %>% 
  arrange(iso3c) -> d5.poorest.latest
d5.poorest.latest %>% kable

#Poorest regions in 2000, or earliest avail. 
dat5.tl3[REG_ID %in% d5.poorest.latest$REG_ID,] %>%
  .[Year >= 2000, .SD[which.min(Year)], by=.(iso3c)] %>%
  .[, geo := "poorest.region"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d5.poorest.earliest

#Adding in poorest region in USA (TL2) in 2000 to TL3 regions in 2000
dat5.tl2[REG_ID %in% d5.poorest.latest$REG_ID,] %>%
  .[Year >= 2000, .SD[which.min(Year)], by=.(iso3c)] %>%
  .[, geo := "poorest.region"] %>%
  .[iso3c == "USA", .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  bind_rows(., d5.poorest.earliest) %>% 
  arrange(iso3c) -> d5.poorest.earliest
d5.poorest.earliest %>% kable



#------- Nationl averages, GDP per person employed, matching years above ------- 


#Nat average, match year to latest year in richest/poorest 
dat5.nat[iso3c %in% d5.richest.latest$iso3c,]  %>%
  inner_join(., d5.richest.latest[,.(iso3c, Year)], by = c("iso3c", "Year")) %>%
  .[, geo := "national.av"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d5.nat.latest
d5.nat.latest %>% kable

#Nat average, 2000 or earliest avail.
dat5.nat[iso3c %in% d5.richest.earliest$iso3c,]  %>%
  inner_join(., d5.richest.earliest[,.(iso3c, Year)], by = c("iso3c", "Year")) %>%
  .[, geo := "national.av"] %>%
  .[, .(iso3c, geo, REG_ID, Region, Measure, Year, Value)] %>%
  arrange(iso3c) -> d5.nat.earliest
d5.nat.earliest %>% kable


```


Again, let's bind the data together and plot our new measure. 

```{r}


#------- Bind data together and plot  -------- 

rbindlist(l = list(d5.nat.latest, d5.nat.earliest, 
                   d5.richest.latest, d5.richest.earliest, 
                   d5.poorest.latest, d5.poorest.earliest)) %>%
  left_join(., oecd.names, by = "iso3c") %>%
  dcast(country + iso3c + Year ~ geo, value.var = "Value") %>%
  mutate(indx.poorest.region = poorest.region / national.av * 100, 
         indx.richest.region = richest.region / national.av * 100) -> rich.poor.2
rich.poor.2 %>% kable

#The countries in our original chart
plot.iso <- c("GBR", "DEU", "USA", "FRA", "KOR", "ITA", "JPA", "ESP", "SWE")

#Plot it
rich.poor.2 %>% 
  filter(iso3c %in% plot.iso) %>%
  select(country, Year, poorest=indx.poorest.region, richest=indx.richest.region) %>% 
  melt(id.vars = c("country", "Year"), variable.name = "geo") -> plot.dat.2
plot.dat.2 %>% kable
write_csv(plot.dat.2, "outputs/plot.dat.2.csv")

ggplot(plot.dat.2, aes(x=value, y=1, colour=Year)) + geom_point() + geom_text(aes(label=geo, vjust=-1)) +
  facet_wrap(~country, ncol=1) + theme_minimal() +
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
  theme(legend.position="top") + ggtitle("Richest v poorest regions, GDP per person employed") -> plot2
print(plot2)

```


